---
title: "Manipulating data"
author: "Licheng Zhang"
date: "1/9/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Data manipulation using dplyr and tidyr

Bracket subsetting is handy, but it can be cumbersome and difficult to read, especially for complicated operations. Enter **dplyr**. **dplyr** is a package for helping with tabular data manipulation. It pairs nicely with **tidyr** which enables you to swiftly convert between different data formats for plotting and analysis.

The **tidyverse** package is an “umbrella-package” that installs **tidyr**, **dplyr**, and several other useful packages for data analysis, such as **ggplot2**, **tibble**, etc.

The **tidyverse** package tries to address 3 common issues that arise when doing data analysis in R:

- The results from a base R function sometimes depend on the type of data.
- R expressions are used in a non standard way, which can be confusing for new learners.
- The existence of hidden arguments having default operations that new learners are not aware of.


```{r}
library(tidyverse)
```


```{r}
surveys <- read_csv('data_raw/survey_data.csv')

# inspect the data
str(surveys)

```


Next, we’re going to learn some of the most common **dplyr** functions:

- select(): subset columns
- filter(): subset rows on conditions
- mutate(): create new columns by using information from other columns
- group_by() and summarize(): create summary statistics on grouped data
- arrange(): sort results
- count(): count discrete values

## Selecting columns and filtering rows
```{r}
select(surveys, plot_id, species_id, weight)

select(surveys, -record_id, -species_id)

filter(surveys, year==1995)
```

## Pipes
what if you want to select and filter at the same time? There are three ways to do this: 

- use intermediate steps
- nested functions
- pipes

```{r}
# use intermediate steps
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)


# nested functions
surveys_sml <- select(filter(surveys, weight < 5), species_id, sex, weight)


# pipes, ctrl+shift+m in PC, cmd+shift+m in mac
surveys_sml <- surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)
surveys_sml

```


Exercise
```{r}
# using pipes , subset the surveys data to include animals collected before 1995 and retain only the columns year, sex and weight



```

Mutate = create new columns
```{r}
# create a new column of weight in kg
surveys %>% 
  mutate(weight_kg = weight / 1000)  # if you don't save it as another object, then it only shows up in the output

# also create a second new column based on the first new column within the same call of mutate()
surveys %>% 
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)

# just check first few rows
surveys %>% 
  mutate(weigt_kg = weight / 1000) %>% 
  head()


# remove NA
surveys %>% 
  filter(!is.na(weight)) %>% 
  mutate(weight_kg = weight / 1000) %>% 
  head()

```

Exercise
```{r}
# Create a new data frame from the surveys data that meets the following criteria: contains only the species_id column and a new column called hindfoot_cm containing the hindfoot_length values (currently in mm) converted to centimeters. In this hindfoot_cm column, there are no NAs and all values are less than 3.

# Hint: think about how the commands should be ordered to produce this data frame!


```



Split-apply-combine data analysis and the summerize() function

Many data analysis tasks can be approached using the split-apply-combine paradigm: 

- split the data into groups
- apply some analysis to each group -
- then combine the results. 

Key function of dplyr for this workflow are group_by() and summarize().

```{r}
# compute mean weight by sex
surveys %>% 
  group_by(sex) %>% 
  summarise(mean_weight = mean(weight, na.rm = TRUE))


# group by multiple columns
surveys %>% 
  group_by(sex, species_id) %>% 
  summarise(mean_weight = mean(weight, na.rm = TRUE)) %>% 
  tail()

# remove missing values first
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight)) %>% 
  tail()


# display more data using print() function
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight)) %>% 
  print(n=10)
```

once the data are grouped, we can summarize multiple variables at the same time. 
```{r}
# add a column indicating the minimum weight for each species for each sex
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) 


# sort on min_weight 
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>% 
  arrange(min_weight)

# sort in descending order
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>% 
  arrange(desc(min_weight))

```


Counting
```{r}
# count the number of rows of data for each sex
surveys %>% 
  count(sex)

# the below is equivalent to above 
surveys %>% 
  group_by(sex) %>% 
  summarize(count = n())

# for convenience
surveys %>% 
  count(sex, sort = TRUE)

# count combination of factors
surveys %>% 
  count(sex, species)

# in addition, we can sort the output 
surveys %>% 
  count(sex, species) %>% 
  arrange(species, desc(n))

```


Exercise
```{r}
# 1. How many animals were caught in each plot_type surveyed?


# 2. Use group_by() and summarize() to find the mean, min, and max hindfoot length for each species (using species_id). Also add the number of observations (hint: see ?n).


# 3. What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and weight.


```



Reshaping with gather and spread

Spreading
```{r}
surveys_gw <- surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(plot_id, genus) %>% 
  summarize(mean_weight = mean(weight))

str(surveys_gw)
head(surveys_gw)

surveys_spread <- surveys_gw %>% 
  spread(key = genus, value = mean_weight)

str(surveys_spread)
head(surveys_spread)


```


Gathering
```{r}
surveys_gather <- surveys_spread %>% 
  gather(key = 'genus', value = 'mean_weight', -plot_id)


str(surveys_gather)

surveys_spread %>% 
  gather(key = 'genus', value = 'mean_weight', Baiomys:Spermophilus)

```


Exercise
```{r}
# 1. Spread the surveys data frame with year as columns, plot_id as rows, and the number of genera per plot as the values. You will need to summarize before reshaping, and use the function n_distinct() to get the number of unique genera within a particular chunk of data. It’s a powerful function! See ?n_distinct for more.


# 2. Now take that data frame and gather() it again, so each row is a unique plot_id by year combination.


# 3. The surveys data set has two measurement columns: hindfoot_length and weight. This makes it difficult to do things like look at the relationship between mean values of each measurement per year in different plot types. Let’s walk through a common solution for this type of problem. First, use gather() to create a dataset where we have a key column called measurement and a value column that takes on the value of either hindfoot_length or weight. Hint: You’ll need to specify which columns are being gathered.


# 4. With this new data set, calculate the average of each measurement in each year for each different plot_type. Then spread() them into a data set with a column for hindfoot_length and weight. Hint: You only need to specify the key and value columns for spread().




```




# Exporting data

```{r}
# removing observations of animals for which weight and hindfoot_length are missing, or the sex has not been determined
surveys_complete <- surveys %>% 
  filter(!is.na(weight),   
         !is.na(hindfoot_length), 
         !is.na(sex))


# Extract the most common species_id
species_counts <- surveys_complete %>%
    count(species_id) %>%
    filter(n >= 50)


# Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)


# save as csv file in data folder
write_csv(surveys_complete, file = 'data/surveys_complete.csv')
```

